Abstractions for Multi-Sorted Substitutions

### 2024-03-13 paper/paper.tex
Update.

  In this subsection, we define the \ACode{\ACon{CKit}}
  abstraction, which provides the \ACode{\AField{\_\&/⋯\_}} operation
  that is necessary for defining composition, and the
  \ACode{\ACon{CTraversal}}, which provides the
  \ACode{\AField{fusion}} lemma for general maps.

  Similarly as it was the case for \ACode{\ACon{Kit}} and \ACode{\ACon{Traversal}},
  the lemma of the \ACode{\ACon{CTraversal}} needs to be proved for
  an abstract \ACode{\ACon{CKit}} and in return defines the
  four concrete \ACode{\ACon{CKit}} instances for us. This allows us to break
  the dependencies between the different \ACode{\AField{fusion}} lemmas,
  similar to how \ACode{\ACon{Kit}} and \ACode{\ACon{Traversal}} allowed
  us to break the dependency between traversals with renamings and substitutions.

  For this purpose, the \ACode{\ACon{CKit}} abstraction, needs
  to satisfy two requirements:
  it needs to provide us with enough machinery to define composition, but
  it also needs to provide us with enough axioms that we are able to instantiate
  the \ACode{\ACon{CTraversal}}. It is defined as followed:
  \LibCode\KCKit
  The \ACode{\AField{\_\&/⋯\_}} operation takes a variable or term
  (according to \ACode{K₁}) and a renaming or substitution
  (according to \ACode{K₂}) and performs the map traversal returning
  a variable or term (according to \ACode{K₁⊔K₂}).
  The third parameter \ACode{K₁⊔K₂} can be seen as a functional
  dependency\cite{DBLP:conf/esop/Jones00} and is determined
  by the choice of \ACode{K₁} and \ACode{K₂}.

  The \ACode{\AField{\&/⋯-⋯}} lemma describes the behavior of
  \ACode{\AField{\_\&/⋯\_}} in terms of \ACode{\AField{\_⋯\_}}, allowing
  subsequent lemmas to make use of the lemmas that we have already proved
  for \ACode{\AField{\_⋯\_}}.

  The \ACode{\AField{\&/⋯-wk-↑}} lemma states that applying a map and then
  weakening is the same as weakening first and then lifting the map over
  the variable introduced by the weakening.

  Similar as for the \ACode{\ACon{Kit}}, we derive further definitions
  from the axioms of a \ACode{\ACon{CKit}}:
  \begin{itemize}
  \item
    We define map composition as motivated by the previous subsection:
    \LibCode\KComposition
  % \item
  %   We show that the \ACode{\AField{\_\&/⋯\_}} operator behaves like
  %   \ACode{\AField{\_\&\_}} when applied to variables:
  %   \LibCode\KCKitAp
  %   % \LibCode\KCKitApProof

  %   % The \ACode{\AField{`/id}} on the left side is from \ACode{K₁⊔K₂}, whereas
  %   % the \ACode{\AField{`/id}} on the right side is from \ACode{K₂}.
  \item
    We show that lifting \ACode{\AField{\_↑\_}} distributes over
    composition:
    \LibCode\KDistLiftCompose
    % \LibCode\KDistLiftComposeProof
  \end{itemize}
  % To make it easier to explicitly specify a \ACode{\ACon{CKit}},
  % we define the notation \ACode{ϕ₁ \ADef{·[} C \ADef{]} ϕ₂}
  % to stand for \ACode{ϕ₁ \ADef{·ₖ} ϕ₂} for some
  % \ACode{\ACon{CKit}} \ACode{C}.
  % \LibCode\KCKitNotation

  We then define a structure,
  which provides the \ACode{\AField{fusion}} lemma for arbitrary
  \ACode{\ACon{CKit}}s:
  \LibCode\KCTraversal

  Similar as for the \ACode{\ACon{Traversal}}, we derive further
  definitions from the axioms of a \ACode{\ACon{CTraversal}}:
  \begin{itemize}
  \item 
    A map \ACode{ϕ} followed by a weakening is equivalent to a
    weakening followed by \ACode{ϕ} that has been lifted over the weakened
    variable:
    \LibCode\KCommLiftWeakenTraverse

    % A map \ACode{ϕ} followed by a weakening is equivalent to a
    % weakening followed by \ACode{ϕ} that has been lifted over the weakened
    % variable:
    % \LibCode\KCommLiftWeaken
    % % \LibCode\KCommLiftWeakenProof
    % By applying \ACode{\AField{fusion}} to this lemma, we get the same
    % statement for multiple map applications:
    % \LibCode\KCommLiftWeakenTraverse
    % % \LibCode\KCommLiftWeakenTraverseProof
  \item 
    A weakening followed by a singleton substitution cancel each other out:
    \LibCode\KWeakenCancelsSingleTraverse

    % A weakening followed by a singleton substitution cancel each other out:
    % \LibCode\KWeakenCancelsSingle
    % % \LibCode\KWeakenCancelsSingleProof
    % By applying \ACode{\AField{fusion}} and \ACode{\AField{⋯-id}} to
    % this lemma, we get the same statement for multiple map applications:
    % \LibCode\KWeakenCancelsSingleTraverse
    % % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton map can be swapped with any map \ACode{ϕ}:
    \LibCode\KDistLiftSingleTraverse
    % A singleton map can be swapped with any map \ACode{ϕ}:
    % \LibCode\KDistLiftSingle
    % % \LibCode\KDistLiftSingleProof
    % By applying \ACode{\AField{fusion}} to this lemma, we get the
    % same statement for multiple map applications:
    % \LibCode\KDistLiftSingleTraverse
    % % \LibCode\KDistLiftSingleTraverseProof
  \item 
    We define \ACode{\ACon{CKit}} instances for each of the four
    combinations between renamings and substitutions:
    \LibCode\KCKitInstances
    Interestingly, the \ACode{\ADef{Cₛ}} instance has another
    \ACode{\ADef{CKit}} parameter.
    This allows us to first use \ACode{\ADef{Cᵣ}} as an argument,
    which yields the \ACode{\ADef{CKit}} between substitutions
    and renamings, and then use this resulting \ACode{\ADef{CKit}}
    as an argument for the case with two substitutions.
    As shown below, Agda's instance resolution is up for the task:

    % To get the \ACode{\ADef{CKit}} between substitutions and
    % renamings, the argument \ACode{\ADef{Cᵣ}} is used.
    % The resulting \ACode{\ADef{CKit}} is then again used as an argument
    % for \ACode{\ADef{Cₛ}} to yield the \ACode{\ADef{CKit}}
    % between two substitutions. As shown below, Agda's instance
    % resolution is up for the task:
    \begin{minipage}[t]{0.5\linewidth}
    \LibCode\KCKitInstancesConcreteI
    \end{minipage}
    \begin{minipage}[t]{0.5\linewidth}
    \LibCode\KCKitInstancesConcreteII
    \end{minipage}

  \end{itemize}

@@

  \AppCode\FPreserve
  The type of \ACode{\ADef{\_⊢⋯\_}} is the same as in the record definition and hence omitted.
  The interesting parts of the proof are:
  \begin{itemize}
  \item
    There is a strong similarity to the instantiation of
    map traversal \ACode{\AField{\_⋯\_}}. Where \ACode{\AField{\_⋯\_}}
    used \ACode{\AField{`/id}} or \ACode{\AField{\_↑\_}}, we now use their
    preservation lemmas \ACode{\AField{⊢`/id}} or \ACode{\AField{\_∋↑/⊢↑\_}}.
  \item
    The lambda typing constructor \ACode{\ACon{⊢λ}} weakens the type \ACode{t₂} to
    shield it from expression-substitution, requiring us to use the
    \ACode{\AField{⋯-↑-wk}} lemma.
  \item
    The type application constructor \ACode{\ACon{⊢∙}} substitutes \ACode{t₁} into \ACode{t₂},
    requiring us to use \ACode{\AField{dist-↑-⦅⦆}}.
  \end{itemize}
  % Finally, we instantiate the \ACode{\ACon{TTraversal}} structure:
  % \AppCode\FTTraversal

>>>>

  % \subsection{An Abstraction for Composition}
  % \label{sec:composition:ckit}

  % In this subsection, we define the \ACode{\ACon{CKit}}
  % abstraction, which provides the \ACode{\AField{\_\&/⋯\_}} operation
  % that is necessary for defining composition, and the
  % \ACode{\ACon{CTraversal}}, which provides the
  % \ACode{\AField{fusion}} lemma for general maps.

  % Similarly as it was the case for \ACode{\ACon{Kit}} and \ACode{\ACon{Traversal}},
  % the lemma of the \ACode{\ACon{CTraversal}} needs to be proved for
  % an abstract \ACode{\ACon{CKit}} and in return defines the
  % four concrete \ACode{\ACon{CKit}} instances for us. This allows us to break
  % the dependencies between the different \ACode{\AField{fusion}} lemmas,
  % similar to how \ACode{\ACon{Kit}} and \ACode{\ACon{Traversal}} allowed
  % us to break the dependency between traversals with renamings and substitutions.

  % For this purpose, the \ACode{\ACon{CKit}} abstraction, needs
  % to satisfy two requirements:
  % it needs to provide us with enough machinery to define composition, but
  % it also needs to provide us with enough axioms that we are able to instantiate
  % the \ACode{\ACon{CTraversal}}. It is defined as followed:
  % \LibCode\KCKit
  % The \ACode{\AField{\_\&/⋯\_}} operation takes a variable or term
  % (according to \ACode{K₁}) and a renaming or substitution
  % (according to \ACode{K₂}) and performs the map traversal returning
  % a variable or term (according to \ACode{K₁⊔K₂}).
  % The third parameter \ACode{K₁⊔K₂} can be seen as a functional
  % dependency\cite{DBLP:conf/esop/Jones00} and is determined
  % by the choice of \ACode{K₁} and \ACode{K₂}.

  % The \ACode{\AField{\&/⋯-⋯}} lemma describes the behavior of
  % \ACode{\AField{\_\&/⋯\_}} in terms of \ACode{\AField{\_⋯\_}}, allowing
  % subsequent lemmas to make use of the lemmas that we have already proved
  % for \ACode{\AField{\_⋯\_}}.

  % The \ACode{\AField{\&/⋯-wk-↑}} lemma states that applying a map and then
  % weakening is the same as weakening first and then lifting the map over
  % the variable introduced by the weakening.

  % Similar as for the \ACode{\ACon{Kit}}, we derive further definitions
  % from the axioms of a \ACode{\ACon{CKit}}:
  % \begin{itemize}
  % \item
  %   We define map composition as motivated by the previous subsection:
  %   \LibCode\KComposition
  % % \item
  % %   We show that the \ACode{\AField{\_\&/⋯\_}} operator behaves like
  % %   \ACode{\AField{\_\&\_}} when applied to variables:
  % %   \LibCode\KCKitAp
  % %   % \LibCode\KCKitApProof

  % %   % The \ACode{\AField{`/id}} on the left side is from \ACode{K₁⊔K₂}, whereas
  % %   % the \ACode{\AField{`/id}} on the right side is from \ACode{K₂}.
  % \item
  %   We show that lifting \ACode{\AField{\_↑\_}} distributes over
  %   composition:
  %   \LibCode\KDistLiftCompose
  %   % \LibCode\KDistLiftComposeProof
  % \end{itemize}
  % % To make it easier to explicitly specify a \ACode{\ACon{CKit}},
  % % we define the notation \ACode{ϕ₁ \ADef{·[} C \ADef{]} ϕ₂}
  % % to stand for \ACode{ϕ₁ \ADef{·ₖ} ϕ₂} for some
  % % \ACode{\ACon{CKit}} \ACode{C}.
  % % \LibCode\KCKitNotation

  % We then define a structure,
  % which provides the \ACode{\AField{fusion}} lemma for arbitrary
  % \ACode{\ACon{CKit}}s:
  % \LibCode\KCTraversal

  % Similar as for the \ACode{\ACon{Traversal}}, we derive further
  % definitions from the axioms of a \ACode{\ACon{CTraversal}}:
  % \begin{itemize}
  % \item 
  %   A map \ACode{ϕ} followed by a weakening is equivalent to a
  %   weakening followed by \ACode{ϕ} that has been lifted over the weakened
  %   variable:
  %   \LibCode\KCommLiftWeakenTraverse

  %   % A map \ACode{ϕ} followed by a weakening is equivalent to a
  %   % weakening followed by \ACode{ϕ} that has been lifted over the weakened
  %   % variable:
  %   % \LibCode\KCommLiftWeaken
  %   % % \LibCode\KCommLiftWeakenProof
  %   % By applying \ACode{\AField{fusion}} to this lemma, we get the same
  %   % statement for multiple map applications:
  %   % \LibCode\KCommLiftWeakenTraverse
  %   % % \LibCode\KCommLiftWeakenTraverseProof
  % \item 
  %   A weakening followed by a singleton substitution cancel each other out:
  %   \LibCode\KWeakenCancelsSingleTraverse

  %   % A weakening followed by a singleton substitution cancel each other out:
  %   % \LibCode\KWeakenCancelsSingle
  %   % % \LibCode\KWeakenCancelsSingleProof
  %   % By applying \ACode{\AField{fusion}} and \ACode{\AField{⋯-id}} to
  %   % this lemma, we get the same statement for multiple map applications:
  %   % \LibCode\KWeakenCancelsSingleTraverse
  %   % % \LibCode\KWeakenCancelsSingleTraverseProof
  % \item 
  %   A singleton map can be swapped with any map \ACode{ϕ}:
  %   \LibCode\KDistLiftSingleTraverse
  %   % A singleton map can be swapped with any map \ACode{ϕ}:
  %   % \LibCode\KDistLiftSingle
  %   % % \LibCode\KDistLiftSingleProof
  %   % By applying \ACode{\AField{fusion}} to this lemma, we get the
  %   % same statement for multiple map applications:
  %   % \LibCode\KDistLiftSingleTraverse
  %   % % \LibCode\KDistLiftSingleTraverseProof
  % \item 
  %   We define \ACode{\ACon{CKit}} instances for each of the four
  %   combinations between renamings and substitutions:
  %   \LibCode\KCKitInstances
  %   Interestingly, the \ACode{\ADef{Cₛ}} instance has another
  %   \ACode{\ADef{CKit}} parameter.
  %   This allows us to first use \ACode{\ADef{Cᵣ}} as an argument,
  %   which yields the \ACode{\ADef{CKit}} between substitutions
  %   and renamings, and then use this resulting \ACode{\ADef{CKit}}
  %   as an argument for the case with two substitutions.
  %   As shown below, Agda's instance resolution is up for the task:

  %   % To get the \ACode{\ADef{CKit}} between substitutions and
  %   % renamings, the argument \ACode{\ADef{Cᵣ}} is used.
  %   % The resulting \ACode{\ADef{CKit}} is then again used as an argument
  %   % for \ACode{\ADef{Cₛ}} to yield the \ACode{\ADef{CKit}}
  %   % between two substitutions. As shown below, Agda's instance
  %   % resolution is up for the task:
  %   \begin{minipage}[t]{0.5\linewidth}
  %   \LibCode\KCKitInstancesConcreteI
  %   \end{minipage}
  %   \begin{minipage}[t]{0.5\linewidth}
  %   \LibCode\KCKitInstancesConcreteII
  %   \end{minipage}

  % \end{itemize}

@@

  \AppCode\FPreserve
  The type of \ACode{\ADef{\_⊢⋯\_}} is the same as in the record definition and hence omitted.
  The interesting parts of the proof are:
  \begin{itemize}
  \item
    There is a strong similarity to the instantiation of
    map traversal \ACode{\AField{\_⋯\_}}. Where \ACode{\AField{\_⋯\_}}
    used \ACode{\AField{`/id}} or \ACode{\AField{\_↑\_}}, we now use their
    preservation lemmas \ACode{\AField{⊢`/id}} or \ACode{\AField{\_⊢↑\_}}.
  \item
    The lambda typing constructor \ACode{\ACon{⊢λ}} weakens the type \ACode{t₂} to
    shield it from expression-substitution, requiring us to use the
    \ACode{\AField{⋯-↑-wk}} lemma.
  \item
    The type application constructor \ACode{\ACon{⊢∙}} substitutes \ACode{t₁} into \ACode{t₂},
    requiring us to use \ACode{\AField{dist-↑-⦅⦆}}.
  \end{itemize}
  % Finally, we instantiate the \ACode{\ACon{TTraversal}} structure:
  % \AppCode\FTTraversal
